---
title: "Business Intelligence Project"
author: "<Champions>"
date: "<15/10/2023>"
output:
  github_document: 
    toc: yes
    toc_depth: 4
    fig_width: 6
    fig_height: 4
    df_print: default
editor_options:
  chunk_output_type: console
---

# Student Details

|                                              |                  |
|----------------------------------------------|------------------|
| **Student ID Number**                        | 126761           |
|                                              | 134111           |
|                                              | 133996           |
|                                              | 127707           |
|                                              | 135859           |
| **Student Name**                             | Virginia Wanjiru |
|                                              | Immaculate Haayo |
|                                              | Trevor Ngugi     |
|                                              | Clarice Muthoni  |
|                                              | Pauline Wairimu  |
| **BBIT 4.2 Group**                           | B                |
| **BI Project Group Name/ID (if applicable)** | Champions        |

# Setup Chunk

**Note:** the following KnitR options have been set as the global defaults: <BR> `knitr::opts_chunk$set(echo = TRUE, warning = FALSE, eval = TRUE, collapse = FALSE, tidy = TRUE)`.

More KnitR options are documented here <https://bookdown.org/yihui/rmarkdown-cookbook/chunk-options.html> and here <https://yihui.org/knitr/options/>.

```{r setup, include=FALSE}
library(formatR)
knitr::opts_chunk$set(
  warning = FALSE,
  collapse = FALSE
)


# STEP 1. Install and Load the Required Packages ----
## mlbench ----
if (require("mlbench")) {
  require("mlbench")
} else {
  install.packages("mlbench", dependencies = TRUE,
                   repos = "https://cloud.r-project.org")
}

## readr ----
if (require("readr")) {
  require("readr")
} else {
  install.packages("readr", dependencies = TRUE,
                   repos = "https://cloud.r-project.org")
}

## caret ----
if (require("caret")) {
  require("caret")
} else {
  install.packages("caret", dependencies = TRUE,
                   repos = "https://cloud.r-project.org")
}

## e1071 ----
if (require("e1071")) {
  require("e1071")
} else {
  install.packages("e1071", dependencies = TRUE,
                   repos = "https://cloud.r-project.org")
}

## factoextra ----
if (require("factoextra")) {
  require("factoextra")
} else {
  install.packages("factoextra", dependencies = TRUE,
                   repos = "https://cloud.r-project.org")
}

## FactoMineR ----
if (require("FactoMineR")) {
  require("FactoMineR")
} else {
  install.packages("FactoMineR", dependencies = TRUE,
                   repos = "https://cloud.r-project.org")
}
```

# Understanding the Structure of Data     

## Loading the Dataset

### Source:

The dataset that was used can be downloaded here: <https://drive.google.com/drive/folders/1-BGEhfOwquXF6KKXwcvrx7WuZXuqmW9q?usp=sharing>

### Reference:

*\
Refer to the APA 7th edition manual for rules on how to cite datasets: <https://apastyle.apa.org/style-grammar-guidelines/references/examples/data-set-references>*

```{r Dataset Loader}
library(readr)
StudentPerformanceDataset <- read_csv("../markdown/StudentPerformanceDataset.csv")

```
 # Applying the Scale Data Transform
```{r Apply Scale Data Transform}


# We use the "preProcess()" function in the caret package

### The Scale Basic Transform on the Student Performance Dataset ----
# BEFORE
class(StudentPerformanceDataset)
summary(StudentPerformanceDataset)
colnames(StudentPerformanceDataset)
StudentPerformanceDataset <- as.data.frame(StudentPerformanceDataset)


hist(StudentPerformanceDataset[,2], main = names(StudentPerformanceDataset)[2])
hist(StudentPerformanceDataset[,3], main = names(StudentPerformanceDataset)[3])
hist(StudentPerformanceDataset[,4], main = names(StudentPerformanceDataset)[5])
hist(StudentPerformanceDataset[,5], main = names(StudentPerformanceDataset)[6])
hist(StudentPerformanceDataset[,6], main = names(StudentPerformanceDataset)[7])
hist(StudentPerformanceDataset[,7], main = names(StudentPerformanceDataset)[8])
hist(StudentPerformanceDataset[,8], main = names(StudentPerformanceDataset)[9])
hist(StudentPerformanceDataset[,9], main = names(StudentPerformanceDataset)[10])
hist(StudentPerformanceDataset[,10], main = names(StudentPerformanceDataset)[11])
hist(StudentPerformanceDataset[,11], main = names(StudentPerformanceDataset)[12])
hist(StudentPerformanceDataset[,12], main = names(StudentPerformanceDataset)[13])
hist(StudentPerformanceDataset[,13], main = names(StudentPerformanceDataset)[14])

student_performance_scale_transform <- as.data.frame(StudentPerformanceDataset)  # Convert to data frame if needed
preprocessed_data <- preProcess(StudentPerformanceDataset)

model_of_the_transform <- preProcess(StudentPerformanceDataset, method = c("scale"))
print(model_of_the_transform)
student_performance_scale_transform <- predict(model_of_the_transform,
                                               StudentPerformanceDataset)
# AFTER
summary(student_performance_scale_transform)


student_performance_scale_transform <- as.data.frame(student_performance_scale_transform)  # Convert to data frame if needed
preprocessed_data <- preProcess(student_performance_scale_transform)



hist(student_performance_scale_transform[,2],
     main = names(student_performance_scale_transform)[2])

hist(student_performance_scale_transform[,3],
     main = names(student_performance_scale_transform)[3])
hist(student_performance_scale_transform[,5],
     main = names(student_performance_scale_transform)[5])
hist(student_performance_scale_transform[,6],
     main = names(student_performance_scale_transform)[6])
hist(student_performance_scale_transform[,7],
     main = names(student_performance_scale_transform)[7])
hist(student_performance_scale_transform[,8],
     main = names(student_performance_scale_transform)[8])
hist(student_performance_scale_transform[,9],
     main = names(student_performance_scale_transform)[9])
hist(student_performance_scale_transform[,10],
     main = names(student_performance_scale_transform)[10])
hist(student_performance_scale_transform[,11],
     main = names(student_performance_scale_transform)[11])
hist(student_performance_scale_transform[,12],
     main = names(student_performance_scale_transform)[12])
hist(student_performance_scale_transform[,13],
     main = names(student_performance_scale_transform)[13])
hist(student_performance_scale_transform[,14],
     main = names(student_performance_scale_transform)[14])


```

# Applying the Center Data Transform 

```{r Apply Center Data Transform}
### The Centre Basic Transform on the Student Performance Dataset ----
# BEFORE
summary(StudentPerformanceDataset)

boxplot(StudentPerformanceDataset[,2], main = names(StudentPerformanceDataset)[2])
boxplot(StudentPerformanceDataset[,4], main = names(StudentPerformanceDataset)[3])
boxplot(StudentPerformanceDataset[,5], main = names(StudentPerformanceDataset)[5])
boxplot(StudentPerformanceDataset[,6], main = names(StudentPerformanceDataset)[6])
boxplot(StudentPerformanceDataset[,7], main = names(StudentPerformanceDataset)[7])
boxplot(StudentPerformanceDataset[,8], main = names(StudentPerformanceDataset)[8])
boxplot(StudentPerformanceDataset[,9], main = names(StudentPerformanceDataset)[9])
boxplot(StudentPerformanceDataset[,10], main = names(StudentPerformanceDataset)[10])
boxplot(StudentPerformanceDataset[,11], main = names(StudentPerformanceDataset)[11])
boxplot(StudentPerformanceDataset[,12], main = names(StudentPerformanceDataset)[12])
boxplot(StudentPerformanceDataset[,13], main = names(StudentPerformanceDataset)[13])
boxplot(StudentPerformanceDataset[,14], main = names(StudentPerformanceDataset)[14])

model_of_the_transform <- preProcess(StudentPerformanceDataset, method = c("center"))
print(model_of_the_transform)
student_performance_center_transform <- predict(model_of_the_transform, # nolint
                                                StudentPerformanceDataset)

# AFTER
summary(student_performance_center_transform)

boxplot(student_performance_center_transform[,2],
        main = names(student_performance_center_transform)[2])
boxplot(student_performance_center_transform[,3],
        main = names(student_performance_center_transform)[3])
boxplot(student_performance_center_transform[,5],
        main = names(student_performance_center_transform)[5])
boxplot(student_performance_center_transform[,6],
        main = names(student_performance_center_transform)[6])
boxplot(student_performance_center_transform[,7],
        main = names(student_performance_center_transform)[7])
boxplot(student_performance_center_transform[,8],
        main = names(student_performance_center_transform)[8])
boxplot(student_performance_center_transform[,9],
        main = names(student_performance_center_transform)[9])
boxplot(student_performance_center_transform[,10],
        main = names(student_performance_center_transform)[10])
boxplot(student_performance_center_transform[,11],
        main = names(student_performance_center_transform)[11])
boxplot(student_performance_center_transform[,12],
        main = names(student_performance_center_transform)[12])
boxplot(student_performance_center_transform[,13],
        main = names(student_performance_center_transform)[13])
boxplot(student_performance_center_transform[,14],
        main = names(student_performance_center_transform)[14])


```
# Standardization of  the Student Performance Dataset
```{r Standardize Data Transform}
### The Standardize Basic Transform on the Student Performance Dataset ----
# BEFORE
summary(StudentPerformanceDataset)


model_of_the_transform <- preProcess(StudentPerformanceDataset,
                                     method = c("scale", "center"))
print(model_of_the_transform)
student_performance_standardize_transform <- predict(model_of_the_transform, # nolint
                                                StudentPerformanceDataset)

# AFTER
summary(student_performance_standardize_transform)
sapply(student_performance_standardize_transform, sd)




```
# Normalized the Dataset ensuring the numerical values between 0 and 1 
```{r Normalize Data Transform}
# Normalizing a dataset implies ensuring the numerical data are
# between [0, 1] (inclusive).

### The Normalize Transform on the Student Performance Dataset ----
summary(StudentPerformanceDataset)
model_of_the_transform <- preProcess(StudentPerformanceDataset, method = c("range"))
print(model_of_the_transform)
student_performance_normalize_transform <- predict(model_of_the_transform, # nolint
                                                   StudentPerformanceDataset)
summary(student_performance_normalize_transform)




```
# Applying the Box Cox Data Transform 

```{r Applying Box Cox Data Transform}
### Box-Cox Power Transform on the Student Performance Dataset ----
# BEFORE
summary(StudentPerformanceDataset)
StudentPerformanceDataset <- as.data.frame(StudentPerformanceDataset)


#Calculate the skewness before the Box-Cox transform

sapply(StudentPerformanceDataset[, 7:13], skewness, type = 2)

#Plot a histogram to view the skewness before the Box-Cox transform


hist(StudentPerformanceDataset[, 7], main = names(StudentPerformanceDataset)[7])
hist(StudentPerformanceDataset[, 8], main = names(StudentPerformanceDataset)[8])
hist(StudentPerformanceDataset[, 9], main = names(StudentPerformanceDataset)[9])
hist(StudentPerformanceDataset[, 10], main = names(StudentPerformanceDataset)[10])
hist(StudentPerformanceDataset[, 11], main = names(StudentPerformanceDataset)[11])
hist(StudentPerformanceDataset[, 12], main = names(StudentPerformanceDataset)[12])
hist(StudentPerformanceDataset[, 13], main = names(StudentPerformanceDataset)[13])


StudentPerformanceDataset <- as.data.frame(StudentPerformanceDataset)  # Convert to data frame if needed
preprocessed_data <- preProcess(StudentPerformanceDataset)

model_of_the_transform <- preProcess(StudentPerformanceDataset, method = c("BoxCox"))
print(model_of_the_transform)


student_performance_box_cox_transform <- predict(model_of_the_transform, # nolint
                                            StudentPerformanceDataset)

# AFTER
summary(student_performance_box_cox_transform)
StudentPerformanceDataset <- as.data.frame(StudentPerformanceDataset)


# Calculate the skewness after the Box-Cox transform
sapply(student_performance_box_cox_transform[, 7:13],  skewness, type = 2)

#Plot a histogram to view the skewness after the Box-Cox transform


hist(student_performance_box_cox_transform[, 7],
     main = names(student_performance_box_cox_transform)[7])

hist(student_performance_box_cox_transform[,8],
     main = names(student_performance_box_cox_transform)[8])

hist(student_performance_box_cox_transform[,9],
     main = names(student_performance_box_cox_transform)[9])

hist(student_performance_box_cox_transform[,10],
     main = names(student_performance_box_cox_transform)[10])

hist(student_performance_box_cox_transform[,11],
     main = names(student_performance_box_cox_transform)[11])

hist(student_performance_box_cox_transform[,12],
     main = names(student_performance_box_cox_transform)[12])

hist(student_performance_box_cox_transform[,13],
     main = names(student_performance_box_cox_transform)[13])






```
# We also tried applying the Yeo Johnson Power Transform on the Dataset
```{r Applying Yeo Johnson Power Transform}
### Yeo-Johnson Power Transform on the Boston Housing Dataset ----
# BEFORE
summary(StudentPerformanceDataset)

# Calculate the skewness before the Yeo-Johnson transform
sapply(StudentPerformanceDataset [,7:13],  skewness, type = 2)

StudentPerformanceDataset <- as.data.frame(StudentPerformanceDataset)  # Convert to data frame if needed
preprocessed_data <- preProcess(StudentPerformanceDataset)

# Plot a histogram to view the skewness before the Box-Cox transform
hist(StudentPerformanceDataset[,7], main = names(StudentPerformanceDataset)[7])
hist(StudentPerformanceDataset[,8], main = names(StudentPerformanceDataset)[8])
hist(StudentPerformanceDataset[,9], main = names(StudentPerformanceDataset)[9])
hist(StudentPerformanceDataset[,10], main = names(StudentPerformanceDataset)[10])
hist(StudentPerformanceDataset[,11], main = names(StudentPerformanceDataset)[11])
hist(StudentPerformanceDataset[,12], main = names(StudentPerformanceDataset)[12])
hist(StudentPerformanceDataset[,13], main = names(StudentPerformanceDataset)[13])


model_of_the_transform <- preProcess(StudentPerformanceDataset, method = c("YeoJohnson"))
print(model_of_the_transform)
student_performance_yeo_johnson_transform <- predict(model_of_the_transform, # nolint
                                                     StudentPerformanceDataset  )

# AFTER
summary(student_performance_yeo_johnson_transform, is.numeric)

# Calculate the skewness after the Yeo-Johnson transform
sapply(student_performance_yeo_johnson_transform[,7:13] ,skewness, type = 2)

# Plot a histogram to view the skewness after the Box-Cox transform
hist(student_performance_yeo_johnson_transform[,7],
     main = names(student_performance_yeo_johnson_transform)[7])
hist(student_performance_yeo_johnson_transform[,8],
     main = names(student_performance_yeo_johnson_transform)[8])
hist(student_performance_yeo_johnson_transform[,9],
     main = names(student_performance_yeo_johnson_transform)[9])
hist(student_performance_yeo_johnson_transform[,10],
     main = names(student_performance_yeo_johnson_transform)[10])
hist(student_performance_yeo_johnson_transform[,11],
     main = names(student_performance_yeo_johnson_transform)[11])
hist(student_performance_yeo_johnson_transform[,12],
     main = names(student_performance_yeo_johnson_transform)[12])
hist(student_performance_yeo_johnson_transform[,13],
     main = names(student_performance_yeo_johnson_transform)[13])


### Yeo-Johnson Power Transform on the Student Performance ----
# BEFORE
summary(student_performance_yeo_johnson_transform)

# Calculate the skewness before the Yeo-Johnson transform
sapply(student_performance_yeo_johnson_transform[,7:13],  skewness, type = 2)
sapply(student_performance_yeo_johnson_transform[,7:13], sd)

model_of_the_transform <- preProcess(student_performance_yeo_johnson_transform,
                                     method = c("YeoJohnson"))
print(model_of_the_transform)
student_performance_yeo_johnson_transform <- predict(model_of_the_transform, # nolint
                                                     student_performance_yeo_johnson_transform)

# AFTER
summary(student_performance_yeo_johnson_transform)

# Calculate the skewness after the Yeo-Johnson transform
sapply(student_performance_yeo_johnson_transform[7:13],  skewness, type = 2)
sapply(student_performance_yeo_johnson_transform[7:13], sd)




```


